We study whether augmenting the Adaptive Biasing Force (ABF) method with a
Fisher--Rao (FR) birth--death correction improves the computation of a
reaction-coordinate free energy, and what kind of improvement it can reasonably be
expected to provide. We test this on \emph{three} systems that deliberately probe
different regimes: a two-dimensional double-well \emph{metastability} model with a
quadrature reference; a two-dimensional \emph{entropic-bottleneck} model with a
narrow transverse channel, for which both the reference free energy and the
conditional law $Y\mid X=x\sim\mathcal N(0,1/(\beta\omega(x)^2))$ are analytic;
and a many-body condensed-phase \emph{Weeks--Chandler--Andersen (WCA) dimer} in a
dense solvent, evaluated against a high-accuracy thermodynamic-integration
reference. In each case the FR step reallocates walkers along the reaction
coordinate by killing replicas in over-represented regions and cloning whole
configurations from under-represented regions, nudging the empirical marginal
$\phat_t$ toward a target $q_t$.

The central finding is not a single universal success story, but a regime map. In
the metastability model, where plain ABF already samples reasonably well,
estimated-target FR is a modest accelerator and the oracle diagnostic shows that
target direction can matter. In the cold entropic bottleneck, ABF is starved and
estimated-target FR gives a large gain ($+\EBgainStrong\%$, $\EBcoldWin/20$ seeds),
but the uniform target nearly matches it and the oracle target is worse than ABF;
here density correction and free-energy accuracy are not the same objective. In
the WCA dimer, tuned FR again improves the free energy reliably
($+\WCAtunedGainPct\%$, $\WCAtunedWin/10$ seeds), while the best accuracy occurs
near the edge of an accuracy--diversity tradeoff. Across the cases, FR is best
understood as a marginal-reallocation mechanism whose value depends on how starved
the ABF estimator is, how quickly cloned orthogonal degrees of freedom
re-equilibrate, and how aggressively birth--death erodes ancestor diversity. The
oracle target is therefore a diagnostic control, not a deployable method and not a
universal upper bound.
