\section{Cross-case synthesis}
\label{sec:synthesis}

The three cases were chosen to span a single axis: \emph{how sample-starved ABF
is}. Read together they tell a more coherent story than any one of them does
alone, and they resolve an apparent contradiction in how the FR \emph{target}
behaves.

\subsection{A starvation spectrum}
Order the cases by how badly plain ABF is starved of reaction-coordinate samples.
At the easy end is the metastability model (\cref{sec:case_meta}), where ABF
already converges well; in the middle is the entropic bottleneck
(\cref{sec:case_eb}), whose difficulty we can dial with temperature; at the hard
end is the condensed-phase WCA dimer (\cref{sec:case_wca}), where the dimer must
displace a dense solvent to cross the barrier. \Cref{tab:synthesis} lays the
headline numbers side by side.

\input{tables/synthesis.tex}

\subsection{The gain grows with starvation}
The practical, estimated-target FR gain tracks the starvation axis almost
monotonically. Where ABF is a strong baseline the gain is small and not
unconditional --- $+\EstIntegImpPct\%$ on the integrated error in the
metastability case, with a marginal final-budget gain --- and where ABF is even
less stressed (the hot entropic bottleneck, $\beta\le4$) FR is mildly
\emph{harmful} ($\EBwarmGainBetaFour\%$ at $\beta=4$), because its birth--death
injects resampling noise into an estimator that was not starved to begin with.
Where ABF is starved the gain is large and reliable: $+\EBgainStrong\%$ in the
cold bottleneck ($\EBcoldWin/20$ seeds) and $+\WCAtunedGainPct\%$ on the WCA dimer
($\WCAtunedWin/10$ seeds), the relative gain growing further with simulation
budget and the largest \emph{absolute} error removed in the densest solvent. The entropic-bottleneck temperature sweep is the cleanest evidence
because it moves along the axis \emph{within a single system}: the same model, the
same code, the gain flipping sign from $\beta=4$ to $\beta=8$ as ABF crosses from
well-sampled to starved.

\subsection{Why the target shape matters less than it first seemed}
The cases appear to disagree about the FR \emph{target}. In the easy metastability
case the oracle target is best and the estimated--oracle gap reads as a
``target-quality bottleneck'' (\cref{sec:q1}). In the WCA dimer the oracle is no
better than the estimated or uniform target, and in the entropic bottleneck the
oracle is actually \emph{worse than plain ABF}. These are not contradictory once
the mechanism is separated from the target.

The \emph{dominant} effect of FR is balanced-resampling variance reduction:
killing over-represented and cloning under-represented whole configurations raises
the number of effectively independent force samples per reaction-coordinate bin,
which is exactly what a starved ABF estimator lacks. This effect needs only a
roughly correct \emph{direction} of imbalance, not an accurate free-energy shape
--- which is why the uniform target nearly matches the estimated one in both
starved cases. Target-shape steering is a \emph{secondary} effect that becomes
visible only when the variance-reduction headroom is already exhausted, i.e.\ when
ABF is well-sampled: there the residual error is limited by the target
\emph{direction}, so the oracle (which knows the exact direction) helps. When ABF
is starved the residual error is variance-limited, the exact shape is irrelevant,
and a target that over-commits to a fixed shape --- the oracle, anchored to a
$\Fref$ that the running bias has not yet matched --- can actively fight the bias
and lose to a neutral uniform target. The metastability ``target-quality
bottleneck'' is therefore a regime-specific reading, not a general law; we are
careful not to carry it into the starved regimes.

\subsection{A single failure mechanism: diversity collapse}
Both the safe regimes and the one observed failure are explained by the same
quantity --- ancestor diversity. Aggressive birth--death erodes the effective
number of independent lineages; the gain survives only while the orthogonal
degrees of freedom re-equilibrate fast enough to re-decorrelate the clones. The
WCA dimer reaches the failure boundary (\texttt{fr\_rate}$\ge1$ reverses the
benefit in the rate ladder, and even \texttt{fr\_rate}$=0.5$ becomes harmful at
production budget, ancestor ESS $\to\!20$--$22$) because its dense solvent
re-equilibrates slowly:
clones stay correlated and the estimator degrades. The entropic bottleneck does
\emph{not} reach a failure boundary even at $\gamma=50$ (ESS $\to\EBessLo$) because
its analytic $y$-channel re-equilibrates almost instantly, so even heavy cloning
does not starve the estimator within the budget. Same mechanism, different
orthogonal relaxation times --- which is why a safe FR rate must be calibrated to
how quickly the orthogonal coordinates of \emph{the system at hand} mix.
