\section{Discussion, limitations, and future directions}
\label{sec:discussion}

\subsection{What the three cases establish (Q4)}
The original single-case study could only ask whether FR was ``icing on the
cake'' (a mild accelerator when ABF already works) or ``snow in the winter'' (a
rescue when ABF alone struggles), and explicitly deferred the question to harder
regimes. The two new cases \emph{are} those harder regimes --- lower temperature,
fewer particles, and a condensed phase --- and the answer is now \emph{both}: FR
is icing for easy ABF (metastability; hot bottleneck) and snow for starved ABF
(cold bottleneck; dense WCA dimer), with the transition governed by how
sample-starved ABF is (\cref{sec:synthesis}).

\subsection{Limitations}
We remain deliberately conservative about scope.
\begin{enumerate}
  \item \textbf{The reaction coordinates are low-dimensional.} All three systems
        use a one-dimensional reaction coordinate (a Cartesian component in the
        2D cases, a scalar bond length in WCA). Multi-dimensional or strongly
        curved collective variables, where the geometric mean-force term and the
        conditional sampling are far harder, are untested.
  \item \textbf{The oracle target is not a method.} In every case the oracle uses
        the reference free energy, the unknown being computed, and is reported
        only as a diagnostic control (\cref{rem:oracle}). Its behaviour is read
        only comparatively; notably, it is \emph{not} uniformly best --- it is
        worse than ABF in the entropic bottleneck --- which is itself part of the
        synthesis, not a deployable result.
  \item \textbf{Production estimators are validated surrogates.} The 2D GPU
        backend uses a binned-smoothed mean-force/marginal estimator and a
        vectorised fixed-$N$ birth--death; the WCA backend uses a constrained
        projection with the geometric term. These pass their validation suites
        and agree with the kernel/TI references to within tolerance, but are not
        bit-identical to the original kernel notebook implementation.
  \item \textbf{Metastability conditional diagnostics are not
        configuration-resolved} (\cref{rem:cond_limit}). The entropic-bottleneck
        case closes this gap with an analytic conditional law, but only for that
        model; the WCA conditional fidelity is argued structurally
        (\cref{rem:manybody}) rather than measured per configuration.
  \item \textbf{The entropic barrier is sub-dominant in Case II.} As stated in
        \cref{sec:case_eb}, the bottleneck modulates the transverse width but the
        energetic double well dominates the difficulty; the $\omega_{\mathrm{in}}$
        sweep is honestly flat, and we do not claim a bottleneck-strength scaling.
\end{enumerate}

\subsection{Failure modes}
The analysis points to a single unifying failure mechanism, diversity collapse,
plus target misspecification (\cref{sec:synthesis}). First, an \emph{overly
aggressive mechanism} erodes ancestor diversity faster than the orthogonal
coordinates can re-decorrelate clones; this is realised in the WCA dimer for
\texttt{fr\_rate}$\ge1$ in the rate ladder (the benefit reverses, ancestor ESS
$\to\!20$), with even \texttt{fr\_rate}$=0.5$ turning harmful at production budget,
and is masked in the entropic bottleneck only because its analytic channel
re-equilibrates instantly. The safe FR rate must therefore be calibrated to the
orthogonal relaxation time of the system. Second, a \emph{noisy or biased target}
caps the attainable benefit; the metastability oracle--estimated gap quantifies
the headroom in the easy regime, while the entropic-bottleneck oracle shows that
in a starved regime an over-committed target can do worse than a neutral uniform
one.

\subsection{Future directions}
\begin{itemize}
  \item \textbf{Adaptive, diversity-aware FR strength.} Tie \texttt{fr\_rate}/
        $\gamma$ to a measured ancestor-ESS or orthogonal-relaxation signal, so
        the rate self-limits before diversity collapses --- capturing the
        early-time speed of an aggressive rate without the late-time failure.
  \item \textbf{Higher-dimensional and curved reaction coordinates}, where
        conditional sampling is genuinely hard and the geometric term is
        nontrivial.
  \item \textbf{Configuration-resolved conditional diagnostics} for the many-body
        case, to measure (not just argue) conditional fidelity per hyperparameter.
  \item \textbf{Sharper entropic stressors.} A model in which the entropic barrier
        dominates the energetic one would test whether the bottleneck width, not
        just the temperature, can drive the gain.
\end{itemize}
