\section{Conclusion}
\label{sec:conclusion}

Across three controlled systems with essentially exact references --- a 2D
metastability model, a 2D entropic bottleneck with an analytic conditional law,
and a many-body condensed-phase WCA dimer --- a Fisher--Rao birth--death
correction helps the Adaptive Biasing Force method \emph{in proportion to how
sample-starved ABF is}. When plain ABF already samples well the practical,
estimated-target FR is a modest convergence accelerator and can even be mildly
harmful (the metastability model, $+\EstIntegImpPct\%$ integrated error; the hot
entropic bottleneck, $\EBwarmGainBetaFour\%$). When ABF is starved the same
correction is a large and reliable win: $+\EBgainStrong\%$ in the cold bottleneck
($\EBcoldWin/20$ seeds) and $+\WCAtunedGainPct\%$ on the WCA dimer
($\WCAtunedWin/10$ seeds), with the advantage growing with simulation budget and
solvent density.

The mechanism is the same in all three cases and is best described as
\emph{balanced-resampling variance reduction}: by cloning whole configurations from
under-represented regions of the reaction coordinate, FR raises the number of
effectively independent force samples per bin, which is exactly what a starved ABF
estimator lacks. This explains the study's most counter-intuitive observation ---
that the precise FR \emph{target} matters little in the starved regimes, where a
uniform target nearly matches the estimated one and an oracle target is no better
(WCA) or even worse (entropic bottleneck) than ABF. Target-shape steering is a
secondary effect that surfaces only when the variance-reduction headroom is spent,
i.e.\ when ABF is well-sampled; the ``target-quality bottleneck'' reading of the
easy metastability case does not generalise. The single failure mode is diversity
collapse under too-aggressive birth--death (the WCA dimer's gain erodes past the
\texttt{fr\_rate}$\approx0.1$--$0.2$ sweet spot and reverses outright for
\texttt{fr\_rate}$\ge1$), bounded by how fast the system's orthogonal
coordinates re-equilibrate cloned replicas.

Two caveats persist throughout. The oracle target uses the reference free energy
and is never a deployable method; it is reported only as a diagnostic whose
behaviour --- best in the easy case, worst in the cold bottleneck --- is itself
part of the synthesis. And the practical recommendation is unambiguous: use the
estimated target, at a moderate rate calibrated to the orthogonal relaxation time,
and expect the largest returns precisely where ABF alone struggles most.
